Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-07-05 survey/survey.tex
global pass

We now have in our hands our desired denotational semantics for types.
We want to create a programming language, which we call $\Pi$, such that the
types and type combinators map to $\bot, \top, \uplus, \times$, and
such that we have ground terms whose denotation are all $16$ type
isomorphisms of Fig.~\ref{type-isos}. This is rather
straightforward, as we can simply do this literally. To make the
analogy with commutative semirings stand out even more, we will use
$0, 1, +$, and $\times$ at the type level, and will denote
``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
``constants'' of the language. Note how we have continued with
the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof terms},
but rather than being witnesses of inhabitation, they are witnesses of
equivalences. Thus we get an unexpected programming language design:

@@

\jc{The notation here is somewhat inconsistent with the above,
mainly in the notation for values.}

@@

  \begin{itemize}
  \item For primitive isomophisms \ensuremath{c}, \ensuremath{!~ c} is given by its
    inverse from Figures~\ref{pi-terms} and~\ref{more-pi}.

@@

\[\begin{array}{rclr}
 \sem{0} & = & \idc \\
 \sem{1} & = & \idc \\
 \sem{1+b} & = & \idc \oplus \sem{b} \\
 \sem{(b_1+b_2)+b_3} & = & \assocrp \odot \sem{b_1 + (b_2 + b_3)} \\
 \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp b_1 + b_2 } \\
 \sem{0 \times b_2} & = & \absorbr \\
 \sem{1 \times b_2} & = & \identlt \odot \sem{b_2} \\
 \sem{(b_1\times b_2)\times b_3} & = & \assocrt \odot \sem{b_1\times (b_2\times b_3)} \\
 \sem{(b_1+b_2)\times b_3} & = & \dist \odot \sem{b_1\times b_3+b_2\times b_3} \\
 \end{array}\]
%originally subcode source isomorphisms.tex:598

@@

Combinators of \ensuremath{\Pi } can be written in terms of the operators
described previously or via a graphical language
similar in spirit to those developed for Geometry of Interaction
\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
\cite{BLUTE1996229,selinger-graphical}. Modulo some conventions and shorthand we
describe here, the wiring diagrams are equivalent to the operator
based (syntactic) description of programs.  \ensuremath{\Pi } combinators
expressed in this graphical language look like ``wiring
diagrams''. Values take the form of ``particles'' that flow along the
wires. Computation is expressed by the flow of particles.

@@

\item
The product type \ensuremath{b_1\times b_2} may be represented both as one wire
labeled \ensuremath{b_1\times b_2} or by two parallel wires labeled \ensuremath{b_1} and
\ensuremath{b_2}. Both representations may be used interchangeably.

@@

\item
When representing complex types like \ensuremath{(b_1\times b_2)+b_3} some visual
grouping of the wires may be done to aid readability. The exact type
however will always be clarified by the context of the diagram.

@@

\item
Associativity is entirely skipped in the graphical language. Hence
three parallel wires may be inferred as \ensuremath{b_1\times (b_2\times b_3)} or
\ensuremath{(b_1\times b_2)\times b_3}, based on the context. This is much like handling of
associativity in the graphical representations of categories as well as
that for monoidal categories.

@@

There are two forms of parallel composition -- combinators
  can be combined additively \ensuremath{c_1+ c_2} (shown on the left) or
  multiplicatively \ensuremath{c_1\times c_2} (shown on the right).

@@

\noindent
\textit{Example.} As an example consider the wiring diagram
of the combinator \ensuremath{c} below:
\begin{flalign*}
& \begin{aligned}
c & : & b \times (1+1) \leftrightarrow b + b \\
c & = & \swapt \odot \dist \odot (\identlt + \identlt) \\
\end{aligned} &&
\end{flalign*}

@@

Other than distributity, $\Pi$ combinators simply allow us to shuffle
the components of a stack around as well as shuffle the positions of
components in the plane.
\end{comment}
% \amr{
%   Let’s have a geometric representation of spaces:
% - A + B means putting A next to B
% - A * B means putting A and B on two perpendicular axes and filling
% the square (volume) between them
% All the combinators would indeed have a natural action and if we have
% a few nice pictures it would be great. What do you think? }

@@

Figure~\ref{type-isos} introduces our desired denotational
semantices, and section~\ref{sec:opsem} a direct definition
of an operational semantics. One obvious question arises:
do these correspond?

@@

%%%%%%%%%
\subsection{Examples}

Surprising expressiveness of Pi:

\begin{quote}
  Ed Fredkin pursued the idea that information must be finite in
  density. One day, he announced that things must be even more simple
  than that. He said that he was going to assume that information
  itself is conserved. “You’re out of you mind, Ed.” I
  pronounced. “That’s completely ridiculous. Nothing could happen in
  such a world. There couldn’t even be logical gates. No decisions
  could ever be made.” But when Fredkin gets one of his ideas, he’s
  quite immune to objections like that; indeed, they fuel him with
  energy. Soon he went on to assume that information processing must
  also be reversible — and invented what’s now called the Fredkin
  gate. (Minsky 1999)
\end{quote}

\begin{verbatim}
examples:
  (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
  conditionals
  toffoli
  fredkin
  revisit adder example from Sec 2
  perhaps a larger example as well (from RC 18 paper?)
\end{verbatim}

\label{sec:langRev-examples}

In this section we introduce programming in \ensuremath{\Pi } through
several simple examples. We subsequently discuss the expressiveness of
the language and discuss various properties.

\label{examples}

@@

\paragraph*{Bit Vectors.}
We can represent $n$-bit words using an n-ary product of
\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
using the type \ensuremath{\mathit{bool}\times (\mathit{bool}\times \mathit{bool})}.  We can perform various
operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
three \ensuremath{\mathit{not}} operations:

@@

\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not} \times (\mathit{not} \times \mathit{not})}

@@

\ensuremath{\mathit{reverse} = \swapt \odot (\swapt \times \idc)~ \odot \assocrt}

\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
many others we will see in this thesis, is formed by sequentially
composing several simpler combinators. Instead of presenting the
operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
for presentation reasons) to flatten the tree into a sequence of
reductions as caused by each component. Such a sequence of reductions
is given below:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \swapt & ((v_2, v_3), v_1) \\
 \swapt \times \idc & ((v_3, v_2), v_1) \\
 \assocrt & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

@@

\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool} \times b_1 \leftrightarrow \mathit{bool} \times b_2}

\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc \times c_1) + (\idc \times c_2)) \odot \factor}

@@

The diagram above shows the input value of type \ensuremath{(1+1)\times b_2}
processed by the distribute operator \ensuremath{\dist}, which converts it into
a value of type \ensuremath{(1\times b_2)+(1\times b_2)}. In the
\ensuremath{\mathit{left}} branch, which corresponds to the
case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was
\ensuremath{\mathit{left} ~()}), the combinator~\ensuremath{c_1} is applied to
the value of type~\ensuremath{b_1}. The right
branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.

@@

By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool}\times \mathit{bool}\leftrightarrow \mathit{bool}\times \mathit{bool}} which negates its
second argument if the first argument is \ensuremath{\mathit{true}}. This gate
\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.

If we iterate this construction once more, the resulting combinator
\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool}\times (\mathit{bool}\times \mathit{bool})\leftrightarrow \mathit{bool}\times (\mathit{bool}\times \mathit{bool})}. The
resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
Toffoli gate.

@@

\jc{I am not a big fan of these pictures -- mainly because I can't give any
intuition based on them of why these can be smoothly deformed into one another.
All the ``pictures'' I can come up with [via purely formal reasoning] sit in at least
5 dimensions -- one for each of the wires, and another to give room to maneuver.}
\end{comment}

@@

The constructs in Fig.~\ref{figc} are significantly more subtle, as they
deal with combinators involving $0$, aka an impossibility.  For example,
\[  {(c \otimes \idc_{0}) \odot \absorbl \Leftrightarrow \absorbl \odot \idc_{0}}
\]
(where we have explicitly annotated the types of $\idc$ for increased clarity)
tells us that of the two ways of transforming from $t_1 \times 0$ to $0$, 
namely first doing some arbitrary transformation $c$ from $t_1$ to $t_2$ and
(in parallel) leaving $0$ alone then eliminating $0$, or first eliminating $0$
then doing the identity (at $0$), are equivalent. This is the ``naturality'' of
$\absorbl$. One item to note is the fact that this combinator is not
irreducible, as the $\idc$ on the right can be eliminated. But that is actually
a property visible at an even higher level (which we will not touch in this
paper).  The next $3$ are similarly expressing the naturality of $\absorbr$,
$\factorzl$ and $\factorzr$.

The next combinator, $\absorbr \Leftrightarrow \absorbl$,
is particularly fascinating: while it says something simple ---
that the two obvious ways of transforming $0 \times 0$ into $0$, namely
absorbing either the left or right zero --- it implies something subtle.
A straightforward proof of $\absorbl$ which proceeds by saying that
$0 \times t$ cannot be inhabited because the first member of the pair
cannot, is not in fact equivalent to $\absorbr$ on $0 \times 0$.
However, if we instead define $\absorbl$ to ``transport'' the
putative impossible first member of the pair to its (equally
impossible) output, then these do form equivalent pairs.
The next few in Fig.~\ref{figc} also express how $\absorbr$ and
$\absorbl$ interact with other combinators. As seen previously,
all of these arise as critical pairs. What is much more subtle here
is that the types involved often are asymmetric: they do not have
the same occurences on the left and right. Such cases are
particularly troublesome for finding normal forms \jc{should add
some citations here}.

@@

\jc{The above is, up to uses of assoc, a linear notation for a
commutative diagram. So draw it.}

>>>>

We now have in our hands our desired denotational semantics for types.
We want to create a programming language, which we call $\Pi$, such
that the types and type combinators map to
$\bot, \top, \uplus, \times$, and such that we have ground terms whose
denotation are all $16$ type isomorphisms of
Fig.~\ref{type-isos}. This is rather straightforward, as we can simply
do this literally. To make the analogy with commutative semirings
stand out even more, we will use $0, 1, +$, and $*$ at the type level,
and will denote ``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms}
shows the ``constants'' of the language. Note how we have continued
with the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof
  terms}, but rather than being witnesses of inhabitation, they are
witnesses of equivalences. Thus we get an unexpected programming
language design:

@@

% \jc{The notation here is somewhat inconsistent with the above,
% mainly in the notation for values.}

@@

  \begin{itemize}
  \item For primitive isomorphisms \ensuremath{c}, \ensuremath{!~ c} is given by its
    inverse from Figures~\ref{pi-terms} and~\ref{more-pi}.

@@

\[\begin{array}{rclr}
 \sem{0} & = & \idc \\
 \sem{1} & = & \idc \\
 \sem{1+b} & = & \idc \oplus \sem{b} \\
 \sem{(b_1+b_2)+b_3} & = & \assocrp \odot \sem{b_1 + (b_2 + b_3)} \\
 \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp b_1 + b_2 } \\
 \sem{0 * b_2} & = & \absorbr \\
 \sem{1 * b_2} & = & \identlt \odot \sem{b_2} \\
 \sem{(b_1 * b_2) * b_3} & = & \assocrt \odot \sem{b_1 * (b_2 * b_3)} \\
 \sem{(b_1+b_2) * b_3} & = & \dist \odot \sem{b_1 * b_3+b_2 * b_3} \\
 \end{array}\]
%originally subcode source isomorphisms.tex:598

@@

Combinators of \ensuremath{\Pi } can be written in terms of the
operators described previously or via a graphical language similar in
spirit to those developed for Geometry of Interaction
\cite{DBLP:conf/popl/Mackie95} and string diagrams for category
theory~\cite{BLUTE1996229,selinger-graphical}. Modulo some conventions
and shorthand we describe here, the wiring diagrams are equivalent to
the operator based (syntactic) description of programs.
\ensuremath{\Pi } combinators expressed in this graphical language
look like ``wiring diagrams''. Values take the form of ``particles''
that flow along the wires. Computation is expressed by the flow of
particles.

@@

\item
The product type \ensuremath{b_1 * b_2} may be represented both as one wire
labeled \ensuremath{b_1 * b_2} or by two parallel wires labeled \ensuremath{b_1} and
\ensuremath{b_2}. Both representations may be used interchangeably.

@@

\item
When representing complex types like \ensuremath{(b_1 * b_2)+b_3} some visual
grouping of the wires may be done to aid readability. The exact type
however will always be clarified by the context of the diagram.

@@

\item
Associativity is entirely skipped in the graphical language. Hence
three parallel wires may be inferred as \ensuremath{b_1 * (b_2 * b_3)} or
\ensuremath{(b_1 * b_2) * b_3}, based on the context. This is much like handling of
associativity in the graphical representations of categories as well as
that for monoidal categories.

@@

There are two forms of parallel composition -- combinators
  can be combined additively \ensuremath{c_1+ c_2} (shown on the left) or
  multiplicatively \ensuremath{c_1 * c_2} (shown on the right).

@@

\noindent
\textit{Example.} As an example consider the wiring diagram
of the combinator \ensuremath{c} below:
\begin{flalign*}
& \begin{aligned}
c & : & b * (1+1) \leftrightarrow b + b \\
c & = & \swapt \odot \dist \odot (\identlt + \identlt) \\
\end{aligned} &&
\end{flalign*}

@@

Other than distributivity, $\Pi$ combinators simply allow us to shuffle
the components of a stack around as well as shuffle the positions of
components in the plane.
\end{comment}
% \amr{
%   Let’s have a geometric representation of spaces:
% - A + B means putting A next to B
% - A * B means putting A and B on two perpendicular axes and filling
% the square (volume) between them
% All the combinators would indeed have a natural action and if we have
% a few nice pictures it would be great. What do you think? }

@@

Figure~\ref{type-isos} introduces our desired denotational
semantics, and section~\ref{sec:opsem} a direct definition
of an operational semantics. One obvious question arises:
do these correspond?

@@

%%%%%%%%%
\subsection{Examples}
\label{sec:langRev-examples}
\label{examples}

At first, it is not immediately clear that a programming language
in which information is preserved could model choice. We recall a
quote by Minsky communicating this concern:

\begin{quote}
  Ed Fredkin pursued the idea that information must be finite in
  density. One day, he announced that things must be even more simple
  than that. He said that he was going to assume that information
  itself is conserved. “You’re out of you mind, Ed.” I
  pronounced. “That’s completely ridiculous. Nothing could happen in
  such a world. There couldn’t even be logical gates. No decisions
  could ever be made.” But when Fredkin gets one of his ideas, he’s
  quite immune to objections like that; indeed, they fuel him with
  energy. Soon he went on to assume that information processing must
  also be reversible — and invented what’s now called the Fredkin
  gate~\cite{Hey:1999:FCE:304763}. 
\end{quote}

% \begin{verbatim}
% examples:
%   (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
%   conditionals
%   toffoli
%   fredkin
%   revisit adder example from Sec 2
%   perhaps a larger example as well (from RC 18 paper?)
% \end{verbatim}

We will however show that one can program all logical gates in
$\Pi$. We will start with a few simply examples and then discuss the
expressiveness of the language and discuss various properties.

@@

\paragraph*{Bit Vectors.}
We can represent $n$-bit words using an n-ary product of
\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
using the type \ensuremath{\mathit{bool} * (\mathit{bool} *  \mathit{bool})}.  We can perform various
operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
three \ensuremath{\mathit{not}} operations:

@@

\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  *  (\mathit{not}  *  \mathit{not})}

@@

\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  *  \idc)~ \odot \assocrt}

\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
many others we will see in this thesis, is formed by sequentially
composing several simpler combinators. Instead of presenting the
operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
for presentation reasons) to flatten the tree into a sequence of
reductions as caused by each component. Such a sequence of reductions
is given below:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \swapt & ((v_2, v_3), v_1) \\
 \swapt  *  \idc & ((v_3, v_2), v_1) \\
 \assocrt & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

@@

\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  *  b_1 \leftrightarrow \mathit{bool}  *  b_2}

\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  *  c_1) + (\idc  *  c_2)) \odot \factor}

@@

The diagram above shows the input value of type \ensuremath{(1+1) *  b_2}
processed by the distribute operator \ensuremath{\dist}, which converts it into
a value of type \ensuremath{(1 *  b_2)+(1 *  b_2)}. In the
\ensuremath{\mathit{left}} branch, which corresponds to the
case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was
\ensuremath{\mathit{left} ~()}), the combinator~\ensuremath{c_1} is applied to
the value of type~\ensuremath{b_1}. The right
branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.

@@

By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} *  \mathit{bool}\leftrightarrow \mathit{bool} *  \mathit{bool}} which negates its
second argument if the first argument is \ensuremath{\mathit{true}}. This gate
\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.

If we iterate this construction once more, the resulting combinator
\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} *  (\mathit{bool} *  \mathit{bool})\leftrightarrow \mathit{bool} *  (\mathit{bool} *  \mathit{bool})}. The
resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
Toffoli gate.

@@

% \jc{I am not a big fan of these pictures -- mainly because I can't give any
% intuition based on them of why these can be smoothly deformed into one another.
% All the ``pictures'' I can come up with [via purely formal reasoning] sit in at least
% 5 dimensions -- one for each of the wires, and another to give room to maneuver.}
% \end{comment}

@@

The constructs in Fig.~\ref{figc} are significantly more subtle, as they
deal with combinators involving $0$, aka an impossibility.  For example,
\[  {(c \otimes \idc_{0}) \odot \absorbl \Leftrightarrow \absorbl \odot \idc_{0}}
\]
(where we have explicitly annotated the types of $\idc$ for increased clarity)
tells us that of the two ways of transforming from $t_1  *  0$ to $0$, 
namely first doing some arbitrary transformation $c$ from $t_1$ to $t_2$ and
(in parallel) leaving $0$ alone then eliminating $0$, or first eliminating $0$
then doing the identity (at $0$), are equivalent. This is the ``naturality'' of
$\absorbl$. One item to note is the fact that this combinator is not
irreducible, as the $\idc$ on the right can be eliminated. But that is actually
a property visible at an even higher level (which we will not touch in this
paper).  The next $3$ are similarly expressing the naturality of $\absorbr$,
$\factorzl$ and $\factorzr$.

The next combinator, $\absorbr \Leftrightarrow \absorbl$, is
particularly fascinating: while it says something simple --- that the
two obvious ways of transforming $0 * 0$ into $0$, namely absorbing
either the left or right zero --- it implies something subtle.  A
straightforward proof of $\absorbl$ which proceeds by saying that
$0 * t$ cannot be inhabited because the first member of the pair
cannot, is not in fact equivalent to $\absorbr$ on $0 * 0$.  However,
if we instead define $\absorbl$ to ``transport'' the putative
impossible first member of the pair to its (equally impossible)
output, then these do form equivalent pairs.  The next few in
Fig.~\ref{figc} also express how $\absorbr$ and $\absorbl$ interact
with other combinators. As seen previously, all of these arise as
critical pairs. What is much more subtle here is that the types
involved often are asymmetric: they do not have the same occurrences
on the left and right. Such cases are particularly troublesome for
finding normal forms.

%% \jc{should add some citations here}.

@@

% \jc{The above is, up to uses of assoc, a linear notation for a
% commutative diagram. So draw it.}
